\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-18 chapters/covert_align.tex
Textcites

\ac{eeg} for the CVSA-ERP dataset was recorded using a SynAmps RT amplifier
(Compumedics Neuroscan, Australia) at 2048Hz and 62 Ag/AgCl active electrodes arranged in the
international 10-10 layout fitted to a standard electrode cap (EASYCAP GmbH,
Germany), with electrodes located at AFz and FCz as ground and reference, respectively.
Using electrolyte gel, electrode impedances were brought below 5k$\Omega$.
Electrodes TP9 and TP10, used for off-line re-referencing, were directly
attached to the skin using stickers for better contact.
The power line frequency in Belgium is 50 Hz.
The participant's eye gaze was registered using an EyeLink 1000 Plus eye tracker (SR Research,
Canada) in non-fixation mode.

Participants signed an informed consent form and were seated at a distance of
60 cm from a CRT-emulating monitor (VPixx
Technologies, Canada) operating at a refresh rate of 120Hz, displaying 6
circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
with a radius of 12.28° of visual angle centered on the midpoint of the screen,
conforming to the interface proposed by \textcite{Treder2010}
(\cref{fig:covert-align/interface/interface}).
A hexagonal layout interface with an empty center and a low number of targets
counteracts target crowding, and as long as the subject’s gaze is within the hexagon of
targets, no other target can be between the subject’s gaze and a covertly
attended target.
Targets are full-contrast white and were intensified by scaling them to a
larger size (5.60° of visual angle, see \cref{fig:covert-align/interface/intensification})
instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
peripheral visual field.
Stimuli were presented using PsychoPy (version 2023.1.3)~\cite{Peirce2019}.

@@

\Cref{fig:covert-align/single-trial-roc-auc-diff} shows a comparison of \ac{rocauc}
for all pairs of \ac{tlda}, \ac{cble},
and \ac{wcble} for single-trial classification to investigate the contributions of
\ac{cble} and \ac{wcble} relative to their first-stage classifier \ac{tlda}.
For this evaluation, epochs were rejected when the peak-to-peak amplitude
exceeded 800$\mu V$, and for the CVSA-ERP dataset, if the user's gaze differed by more than
10 degrees of visual angle from the fixation crosshair.
\begin{figure}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:covert-align/single-trial-roc-auc-diff}%
		\includegraphics[width=\linewidth]{figures/covert_align/figure4a.pdf}
	\end{subfigure}

@@

Although it is not immediately clear if \ac{wcble} actively corrects for higher P3 jitter
present in covert and split \ac{vsa} compared to overt \ac{vsa},
we justify our approach similarly to \cite{Hardiansyah2020} by
observing that the increased discrimination performance of a machine-learning
model accounting for jitter forgoes the need for characterizing the underlying
physiological processes, while still objectively quantifying the presence of
jitter between data classes.
Furthermore, some evidence points toward higher P3 jitter as the main
contributing factor.
While \autoref{fig:erps} shows no visible smearing effect in
the shape of the \acp{erp}, a more quantitative analysis on latencies in
\autoref{fig:jitter} indicates the opposite.
To obtain comparable results across \ac{vsa} conditions, \ac{wcble} was
evaluated per session and trained on all combined \ac{vsa} conditions as in
\cref{sec:covert-align/results/cross}.
Since all conditions have the P3 in common, the estimated latencies can be
interpreted as P3 latencies.
Two-sided Wilcoxon rank-sum tests on jitter, expressed as the interquartile
range (IQR) of the estimated latencies of target trials, revealed that overt
\ac{vsa} exhibited significantly lower jitter than all other conditions for
both datasets.\todo{qualitative}
No other significant differences were found.
$p$-values were corrected for the family-wise error rate using Bonferroni
correction.
Additionally, \textcite{Arico2014} prove that P3 jitter does play a
non-negligible role in covert \ac{vsa} by comparing performance between
overt and covert \ac{vsa}, while including or excluding early VEPs.
This analysis showed that the absence of the N1 and other early VEPs is not the
only factor hampering covert \ac{vsa} decoding performance.
However, the large increase in performance for covert \ac{vsa} reported by
\cite{Arico2014}\todo{qualitative} could also
be explained by overfitting on artifacts amplified by aligning, since their
method is not evaluated on unseen data, as opposed to ours.\todo{qualitative}

@@

We found that \ac{cble} did not improve gaze-independent decoding performance
significantly and also did not increase performance over its first-stage
classifier in overt \ac{vsa}, contrary to what was reported by \cite{Mowla2017}.
While they report that \ac{cble} is relatively independent of the
first-stage classifiers evaluated in their work, it is evident here that
applying any given classifier in the \ac{cble} scheme does not necessarily increase
its performance.
In our case, this could be due to characteristics of the tested dataset, e.g.,
the presence of jitter or the generally higher performance of \ac{tlda} compared to the
first-stage classifiers tested by \cite{Mowla2017}, leaving less
performance to be gained.

\cite{Thompson2012} already attempted applying \ac{cble} in an iterative scheme
but did not report any results due to convergence issues.
We mitigated this by combining the robust latency estimation presented in
\cref{sec:robust-latency} with the alignment of both target and non-target
epochs.
Aligning only the target epochs containing the jittered P3 component is prone to
overfitting by aligning non-discriminative noise that is present in both
classes, such as environment noise, oscillatory background rhythms, or
non-modulated VEPs.
If \ac{snr} is low, residual noise varying slightly between classes could dominate
the expected response of the first-stage classifier and subsequently dominate
\ac{wcble} from the start, preventing convergence to a meaningful solution.
Our procedure circumvents this problem by aligning both classes to the time
points where the expected separation between classes is greatest.
This way, noise of which the latencies are estimated in a given iteration will be
perfectly time-locked in all classes in the next
iteration.
The first-stage classifier can then more easily suppress this noise since it is
now clear it is present in both classes and non-discriminative.
This aids the method in converging to a more robust classifier by iteratively
increasing \ac{snr} for both classes and class separation over time.

\cite{Zisk2022} addressed P3 jitter inparticipants \ac{als} by augmenting the
training data once with time-shifted copies based on \ac{cble}-estimated jitter.
While we aim to train the first-stage classifier without the effects of jitter
in its parameters, they do the opposite by intentionally jittering the
training data.
As their focus was on \ac{als} and inter-session stability, they
did not assess how their method interacts with visuospatial attention.
We achieved a similar performance gain with our jitter compensation method, but
argue that our method can cope with more granular latency differences, as
\cite{Zisk2022} augment the data with just one positive and negative time
shift.

>>>>

\ac{eeg} for the CVSA-ERP dataset was recorded using a SynAmps RT amplifier
(Compumedics Neuroscan, Australia) at 2048 Hz and 62 Ag/AgCl active electrodes arranged in the
international 10-10 layout fitted to a standard electrode cap (EASYCAP GmbH,
Germany), with electrodes located at AFz and FCz as ground and reference, respectively.
Using electrolyte gel, electrode impedances were brought below 5k$\Omega$.
Electrodes TP9 and TP10, used for off-line re-referencing, were directly
attached to the skin using stickers for better contact.
The power line frequency in Belgium is 50 Hz.
The participant's eye gaze was registered using an EyeLink 1000 Plus eye tracker (SR Research,
Canada) in non-fixation mode.

Participants signed an informed consent form and were seated at a distance of
60 cm from a CRT-emulating monitor (VPixx
Technologies, Canada) operating at a refresh rate of 120 Hz, displaying 6
circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
with a radius of 12.28° of visual angle centered on the midpoint of the screen,
conforming to the interface proposed by \textcite{Treder2010}
(\cref{fig:covert-align/interface/interface}).
A hexagonal layout interface with an empty center and a low number of targets
counteracts target crowding, and as long as the subject’s gaze is within the hexagon of
targets, no other target can be between the subject’s gaze and a covertly
attended target.
Targets are full-contrast white and were intensified by scaling them to a
larger size (5.60° of visual angle, see \cref{fig:covert-align/interface/intensification})
instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
peripheral visual field.
Stimuli were presented using PsychoPy (version 2023.1.3)~\cite{Peirce2019}.

@@

\Cref{fig:covert-align/single-trial-roc-auc-diff} shows a comparison of \ac{rocauc}
for all pairs of \ac{tlda}, \ac{cble},
and \ac{wcble} for single-trial classification to investigate the contributions of
\ac{cble} and \ac{wcble} relative to their first-stage classifier \ac{tlda}.
For this evaluation, epochs were rejected when the peak-to-peak amplitude
exceeded 800 µV, and for the CVSA-ERP dataset, if the user's gaze differed by more than
10 degrees of visual angle from the fixation crosshair.
\begin{figure}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:covert-align/single-trial-roc-auc-diff}%
		\includegraphics[width=\linewidth]{figures/covert_align/figure4a.pdf}
	\end{subfigure}

@@

Although it is not immediately clear if \ac{wcble} actively corrects for higher P3 jitter
present in covert and split \ac{vsa} compared to overt \ac{vsa},
we justify our approach in a similar manner to \textcite{Hardiansyah2020} by
observing that the increased discrimination performance of a machine-learning
model accounting for jitter forgoes the need for characterizing the underlying
physiological processes, while still objectively quantifying the presence of
jitter between data classes.
Furthermore, some evidence points toward higher P3 jitter as the main
contributing factor.
While \autoref{fig:erps} shows no visible smearing effect in
the shape of the \acp{erp}, a more quantitative analysis on latencies in
\autoref{fig:jitter} indicates the opposite.
To obtain comparable results across \ac{vsa} conditions, \ac{wcble} was
evaluated per session and trained on all combined \ac{vsa} conditions as in
\cref{sec:covert-align/results/cross}.
Since all conditions have the P3 in common, the estimated latencies can be
interpreted as P3 latencies.
Two-sided Wilcoxon rank-sum tests on jitter, expressed as the interquartile
range (IQR) of the estimated latencies of target trials, revealed that overt
\ac{vsa} exhibited significantly lower jitter than all other conditions for
both datasets.\todo{qualitative}
No other significant differences were found.
$p$-values were corrected for the family-wise error rate using Bonferroni
correction.
Additionally, \textcite{Arico2014} prove that P3 jitter does play a
non-negligible role in covert \ac{vsa} by comparing performance between
overt and covert \ac{vsa}, while including or excluding early VEPs.
This analysis showed that the absence of the N1 and other early VEPs is not the
only factor hampering covert \ac{vsa} decoding performance.
However, the large increase in performance for covert \ac{vsa} reported by
\textcite{Arico2014}\todo{qualitative} could also
be explained by overfitting on artifacts amplified by aligning, since their
method is not evaluated on unseen data, as opposed to ours.\todo{qualitative}

@@

We found that \ac{cble} did not improve gaze-independent decoding performance
significantly and also did not increase performance over its first-stage
classifier in overt \ac{vsa}, contrary to what was reported by \textcite{Mowla2017}.
While they report that \ac{cble} is relatively independent of the
first-stage classifiers evaluated in their work, it is evident here that
applying any given classifier in the \ac{cble} scheme does not necessarily increase
its performance.
In our case, this could be due to characteristics of the tested dataset, e.g.,
the presence of jitter or the generally higher performance of \ac{tlda} compared to the
first-stage classifiers tested by \textcite{Mowla2017}, leaving less
performance to be gained.

\textcite{Thompson2012} already attempted applying \ac{cble} in an iterative scheme
but did not report any results due to convergence issues.
We mitigated this by combining the robust latency estimation presented in
\cref{sec:robust-latency} with the alignment of both target and non-target
epochs.
Aligning only the target epochs containing the jittered P3 component is prone to
overfitting by aligning non-discriminative noise that is present in both
classes, such as environment noise, oscillatory background rhythms, or
non-modulated VEPs.
If \ac{snr} is low, residual noise varying slightly between classes could dominate
the expected response of the first-stage classifier and subsequently dominate
\ac{wcble} from the start, preventing convergence to a meaningful solution.
Our procedure circumvents this problem by aligning both classes to the time
points where the expected separation between classes is greatest.
This way, noise of which the latencies are estimated in a given iteration will be
perfectly time-locked in all classes in the next
iteration.
The first-stage classifier can then more easily suppress this noise since it is
now clear it is present in both classes and non-discriminative.
This aids the method in converging to a more robust classifier by iteratively
increasing \ac{snr} for both classes and class separation over time.

\textcite{Zisk2022} addressed P3 jitter inparticipants \ac{als} by augmenting the
training data once with time-shifted copies based on \ac{cble}-estimated jitter.
While we aim to train the first-stage classifier without the effects of jitter
in its parameters, they do the opposite by intentionally jittering the
training data.
As their focus was on \ac{als} and inter-session stability, they
did not assess how their method interacts with visuospatial attention.
We achieved a similar performance gain with our jitter compensation method, but
argue that our method can cope with more granular latency differences, as
\textcite{Zisk2022} augment the data with just one positive and negative time
shift.

### 2024-11-18 chapters/gaze_independence.tex
Textcites

P3 latency generally falls between 350ms and 600ms~\cite{Luck2014}, but this
value is heavily dependent on the subject and the task and can vary from trial
to trial~\cite{Ouyang2017}.
The work of \cite{Arico2014} illustrates that the variation in single-trial P3
latencies is important in gaze-independent decoding and has been hampering covert
\ac{vsa} decoding performance.
We reprised their hypothesis, stating that jitter compensation through latency
estimation and alignment improves
covert \ac{vsa} performance and extended it by developing a decoder.

>>>>

P3 latency generally falls between 350 ms and 600 ms~\cite{Luck2014}, but this
value is heavily dependent on the subject and the task and can vary from trial
to trial~\cite{Ouyang2017}.
The work of \cite{Arico2014} illustrates that the variation in single-trial P3
latencies is important in gaze-independent decoding and has been hampering covert
\ac{vsa} decoding performance.
We reprised their hypothesis, stating that jitter compensation through latency
estimation and alignment improves
covert \ac{vsa} performance and extended it by developing a decoder.
